---
title: "In-class template"
author: "Abby Malech"
format: pdf
---

# 1. Set up

## Packages

```{r}
#| label: packages

# general use and cleaning
library(tidyverse)
library(janitor)
library(here)
library(snakecase)

# wrap axis labels
library(scales)

# reading .xlsx files
library(readxl)

# visualize data using naniar
library(naniar)

# arrange plot using patchwork
library(patchwork)
```

## Data

```{r}
#| label: data

# taxonmic information
taxon_list <- read_csv(here("data", "taxon_list.csv"))

# aquatic invertebrates
aq_ins <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "Aquatic Insects"
)

# site information
sites <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "sites"
)

# water quality
water_quality <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "Water Quality",
  na = c("", "n/a", "N/A", "over", "173+")
)

```

# 2. Cleaning

## Clean `sites` object

Goal: create a new object from sites that only includes NCOS sites

```{r}
#| label: NCOS-sites

# creating new object from sites
ncos_sites <- sites |> 
  # filter to only inclue NCOS quarterly sampling sites
  filter(sector == "NCOS" & sampling_frequency == "Quarterly")
```

## Clean `taxon_list`

Goal: convert character strings in the existing `taxon_list` data frame into snakecase.

```{r}
#| label: cleaning-taxon

# creating new object from taxon_list
taxon_list_clean <- taxon_list |> 
  # converted names using snakecase
  mutate(verbatim_name = to_any_case(verbatim_name,
                                     case = "snake"))
```

## Clean `water_quality`

Goal: create a data frame with median pH, meian D.O. (mg/L), median salinity (ppt) for each site and date

- Create a new object from `water_quality`
- Clean column names using janitor
- Only include surveys from quarterly NCOS sites
- Create a new column called `date_site` to join with later data frames
- Group by date_site
- Calculate median pH, salinity, and dissolved oxygen
- Ungroup data frame
- Filter out any outliers

```{r}
#| label: cleaning-water-quality

# creating new oject from water_quality
water_quality_clean <- water_quality |> 
  # cleaning column names
  clean_names() |>
  # using filter join
  semi_join(ncos_sites, by = "site") |> 
  # uniting by site and date
  unite("date_site", date, site, remove = TRUE) |> 
  # grouping by date site
  group_by(date_site) |> 
  # calculating median ph, salinity, and DO
  summarize(med_ph = median(p_h, na.rm = TRUE),
            med_sal = median(salinity_ppt, na.rm = TRUE),
            med_do = median(dissolved_oxygen_mg_l, na.rm = TRUE)) |> 
  # ungrouping data frame
  ungroup() |> 
  # filtering outliers
  filter(date_site != "2022-08-12_NVBR")
```

## Clean `aq_ins`

Goal: create a long format data frame in which rach row contains the counted individuals in each taxon during each survey. The data frame should include taxonomic information for each organism and the water quality

- create a new object from `aq_ins`
- clean column names
- filter to include only quarterly NCOS sites
- select columns
- filter to only include "filtered beaker" sample
- convert data frame to long format
- group by site, date, and taxon
- calculate average abundance per liter (sum count divide by 7.5 liters)
- create new column called `date_site` to join with water quality
- join data frames
- join with taxon list

```{r}

aq_ins_clean <- aq_ins |> 
  clean_names() |> 
  semi_join(ncos_sites, by = 'site') |> 
  rename(date = date_on_vial) |> 
  select(date, sample_type, site,
         ostracod,
         copepod,
         hemiptera_corixidae_boatman,
         cladocera,
         nematode,
         diptera_ceratopogonidae, 
         annelida_oligochaete,
         diptera_chironomid,
         annelida_polychaete) |> 
  filter(sample_type %in% c("FB250", "FB 250")) |> 
  pivot_longer(
    # columns to make longer
    cols = ostracod:annelida_polychaete,
    # new column for the old column names
    names_to = "taxon",
    # new columns for the values in each of the old columns
    values_to = "abundance"
  ) |> 
  group_by(date, site, taxon) |> 
  summarize(ave_lit = sum(abundance, na.rm = TRUE)/7.5) |> 
  ungroup() |> 
  unite("date_site", date, site, remove = FALSE) |>
  left_join(water_quality, by "date_site") |> 
  left_join(taxon_list_clean, by = c("taxon" = "verbatim_name")) |> 
  mutate(site_full = case_when(
    site == "NVBR" ~ "North of Venoco Bridge",
    site == "NEC" ~ "South of Venoco Bridge",
    site == "NMC" ~ "Main Channel",
    site == "NPB" ~ "South of Phelps Bridge",
    site == "NPB1" ~ ""
  ))




```






