---
title: "Individual Assignment 5"
author: "Abby Malech"
format: 
  pdf:
    include-in-header: 
       text: |
         \usepackage{fvextra}
         \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
         \DefineVerbatimEnvironment{OutputCode}{Verbatim}{breaklines,commandchars=\\\{\}}
date: "2026-05-13"
execute:
  message: false
  warning: false
---

# Packages
```{r}
#| label: load-packages

# general use and customization
library(tidyverse)
library(janitor)
library(here)
library(snakecase)

# wrapping axis labels
library(scales)

# reading in .xlsx files
library(readxl)

# visualizing missing data
library(naniar)

# arranging plots
library(patchwork)
```

# Load Data
```{r}
#| label: load-data

# taxonomic information
taxon_list <- read_csv(here("data", "taxon_list.csv")) 

# aquatic invertebrates
aq_ins <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                    sheet = "Aquatic Insects") 

# site information
sites <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                   sheet = "sites")

# water quality information
water_quality <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                           sheet = "Water Quality",
                           na = c("", "n/a", "N/A", "over", "173+"))
```

# Set Up

```{r}
#| label: in-class-code
#| results: 'hide'
#| fig-show: 'hide'

# creating new object from sites
ncos_sites <- sites |> 
  # filter to only include NCOS sites sampled four times a year
  filter(sector == "NCOS" & sampling_frequency == "Quarterly")

# creating new clean object from taxon_list
taxon_list_clean <- taxon_list |> 
  # converting taxon name to snake case (no spaces or capital letters,
  # only underscores)
  mutate(verbatim_name = to_any_case(verbatim_name, case = "snake"))

water_quality_clean <- water_quality |> 
  # clean column names
  clean_names() |> 
  # filtering join: only include sites in ncos_sites
  semi_join(ncos_sites, by = "site") |> 
  # create new column to join with later data frames
  unite("date_site", date, site, remove = TRUE) |> 
  # group by date_site column
  group_by(date_site) |> 
  # calculate median pH, salinity, DO
  summarize(med_ph = median(p_h, na.rm = TRUE),
            med_sal = median(salinity_ppt, na.rm = TRUE),
            med_do = median(dissolved_oxygen_mg_l, na.rm = TRUE)) |> 
  # ungroup data frame
  ungroup() |> 
  # filter out salinity outliers
  filter(date_site != "2022-08-12_NVBR")

# creating new clean object from aquatic inverts
aq_ins_clean <- aq_ins |> 
  # clean column names
  clean_names() |> 
  # filtering join: only include sites occurring in ncos_sites
  semi_join(ncos_sites, by = c("site")) |> 
  # renaming columns
  rename(date = date_on_vial) |> 
  # select columns of interest
  select(date, sample_type, site,
         ostracod,
         copepod,
         hemiptera_corixidae_boatman,
         cladocera,
         nematode,
         diptera_ceratopogonidae, 
         annelida_oligochaete,
         diptera_chironomid,
         annelida_polychaete) |> 
  # filter to only include "filtered beaker" samples
  filter(sample_type %in% c("FB 250", "FB250")) |> 
  # convert the data frame to long format
  pivot_longer(cols = ostracod:annelida_polychaete,
               names_to = "taxon",
               values_to = "abundance") |> 
  # group by date, site, taxon
  group_by(date, site, taxon) |> 
  # calculate average abundance per liter (sum abundance divided by 7.5 liters)
  summarize(ave_lit = sum(abundance, na.rm = TRUE)/7.5) |> 
  # ungroup data frame
  ungroup() |> 
  # create a new column called `date_site` to join with water quality
  unite("date_site", date, site, remove = FALSE) |> 
  # join with water quality
  left_join(water_quality_clean, by = c("date_site")) |> 
  # join with taxon list
  left_join(taxon_list_clean, by = c("taxon" = "verbatim_name")) |> 
  # add full names of sites
  mutate(site_full = case_when(
    site == "NVBR" ~ "North of Venoco Bridge",
    site == "NEC" ~ "South of Venoco Bridge",
    site == "NMC" ~ "Main Channel",
    site == "NPB" ~ "South of Phelps Bridge",
    site == "NPB1" ~ "North of Phelps Bridge",
    site == "NPB2" ~ "Phelps Road",
    site == "NWP" ~  "West Pond",
    site == "NDC" ~ "Devereux Creek"
  )) |> 
  # setting factor levels for site
  mutate(site_full = fct_reorder(site_full, med_sal, .fun = "median", 
                                 .na_rm = TRUE),
         site_full = fct_rev(site_full))

# creating new object from clean aquatic inverts 
copepods <- aq_ins_clean |> 
  # filter to only include copepods
  filter(taxon == "copepod") |> 
  # log + 1 transform mean abundance per liter
  mutate(log_ave_lit = log(ave_lit + 1)) 

# base layer
copepods <- ggplot(data = copepods,
                   # x-axis: site
                   # y-axis: mean abundance/L
                   mapping = aes(x = site_full,
                                 y = log_ave_lit)) +
  # first layer: points to represent observations
  geom_jitter(height = 0,
              shape = 21,
              width = 0.1) +
  # second layer: points to represent medians
  stat_summary(geom = "point",
               fun = "median",
               size = 4,
               color = "blue") +
  # wrapping x-axis text
  scale_x_discrete(labels = label_wrap(15)) +
  # relabelling x- and y-axis, adding title
  labs(x = "Site",
       y = "log + 1 transformed average abundance/L",
       title = "Copepods") +
  # different theme
  theme_bw()

# base layer
salinity <- ggplot(data = aq_ins_clean,
       # x-axis: site
       # y-axis: mean salinity
       mapping = aes(x = site_full,
                     y = med_sal)) +
  # first layer: salinity measurements at each survey
  geom_jitter(height = 0,
              width = 0.1,
              shape = 21) +
  # wrapping x-axis text
  scale_x_discrete(labels = label_wrap(width = 15)) +
  # second layer: points to represent means
  stat_summary(fun = median,
               color = "red",
               size = 1) +
  # relabelling x- and y-axes
  labs(x = "Site",
       y = "Median salinity (ppt)",
       title = "Site salinity") +
  # different theme
  theme_bw()
```

# 1. Visualizing Differences across sites in major taxon abundance

## a. Plan your figure

- x-axis: site ()
- y-axis: average abundance/L (log transformed)
- geometry to represent average abundance/L: geom_jitter 
- geometry to represent medians:
stat_summary(geompoint)

## b. Wrangle your data

```{r}
#| label: cleaning-aq-ins-clean

# new object from aq_ins_clean
avg_ostracod <- aq_ins_clean |>
# filter to only have ostacod observations
  filter(taxon == "ostracod", rm.na = TRUE) |>
  # log + 1 transform mean abundance per liter
  mutate(log_ave_lit = log(ave_lit + 1))

# sample of ten rows
slice_sample(avg_ostracod, n = 10)
```

## c. Code your figure

```{r}
#| label: plotting-ostracod
#| fig-width: 10
#| fig-height: 8

# creating object and adding ggplot layer using avg_ostacod
ostracod1 <- ggplot(data = avg_ostracod,
# x-axis: site
# y-axis: log trans liter abundance
                    mapping = aes(x = site_full,
                                  y = log_ave_lit)) +
# jitter for individual mean points
  geom_jitter(height = 0, shape = 21, width = 0.1) +
# adding median points
  stat_summary(geom = "point",
               fun = "median",
               size = 4,
               color = "purple") +
# wrapping x-axis text for no overlap
    scale_x_discrete(labels = label_wrap(15)) +
# adding labels
  labs(x = "Site",
       y = "log + 1 transformed average abundance/L",
       title = "Ostracod abundance across NCOS sites") +
# removing space between x-axis and 0
  scale_y_continuous(limits = c(-0.1, 8),
                     expand = c(0, 0)) + 
# changing theme from default
  theme_bw()

ostracod1
```

## d. Create a multipanel figure

```{r}
#| label: multipanel-fig
#| fig-width: 14
#| fig-height: 12

# creating stacked graphs
# salinity on top; then copepods - ostracod underneath
salinity / (copepods | ostracod1)
```

## e. Interpret your figure

The average abundance of Ostracods are relatively low (around 0), with the exception of South of Phelps Bridge and Devereux Creek with a higher abundance of around 2 per liter. Copepod abundance is generally higher but more variable, with the main channel and N. phelps bridge locations relatively low and all other locations having around 2-3 copepods per liter. 

There seems to be a negative correlation between ostracod abundance and site salinity levels, as the site lowers in median salinity levels, ostracod abundance increases on average. There are a few exceptions - namely N. phelps bridge and phelps road that does not follow this trend - but generally it seems that the main channel abundance increases as its salinity drops along with S. Phelps bridge and Devareux creek, which could point to ostracods preferring freshwater over saltier water. 

# 2. Visualizing total abundance as a function of salinity

## a. Plan your figure

- x-axis: median salinity
- y-axis: log +1 avg abundance/L
- color: taxon
- geometry: geom_line(), geom_point()
- facets: facet_wrap(~taxon)

## b. Wrangle your data

```{r}
#| label: clean-data

# creating new object from aq_ins_clean
taxon_sal <- aq_ins_clean |> 
  # mutate to get taxon group -species
  mutate(taxon = case_when(
    str_detect(taxon, "annelida") ~ "Annelida",
    str_detect(taxon, "cladocera") ~ "Cladocera",
    str_detect(taxon, "copepod") ~ "Copepod",
    str_detect(taxon, "diptera") ~ "Diptera",
    str_detect(taxon, "hemiptera") ~ "Hemiptera",
    str_detect(taxon, "nematode") ~ "Nematode",
    str_detect(taxon, "ostracod") ~ "Ostracod"
  )) |> 
  # mutate to get log transformed abundance
  mutate(log_ave_lit = log(ave_lit + 1)) |> 
  # group by taxon
  group_by(taxon) |> 
  # select only necessary columns
  select(taxon, date, log_ave_lit, med_sal) |>
  # remove na values from med_sal
  subset(!is.na(med_sal)) |> 
  # order taxon levels by maximum abundance
  arrange(desc(log_ave_lit)) |> 
  # ungroup dataset
  ungroup()

# displaying 10 random rows
slice_sample(taxon_sal, n = 10)
```

## c. Code your figure

```{r}
#| label: visualization-sal-taxon
#| fig-width: 14
#| fig-height: 12

# create ggplot layer
ggplot(data = taxon_sal,
# x-axis: median salinity
# y-axis: log transformed abundance
# color: by taxon
       mapping = aes(x = med_sal,
                     y = log_ave_lit,
                     color = taxon)) +
# creating panels for each taxon group
  facet_wrap(~taxon, scales = "free") +
# adding line geometries
  geom_line() +
# adding point geometries
  geom_point() +
# adding labels
  labs(x = "Median Salinity (ppt)",
       y = "log +1 transformed average abundance/L",
       title = "Taxon abundance with salinity measurements") +
# adding custom colors
  scale_color_manual(
    values = c("Annelida" = "tan",
               "Cladocera" = "orange",
               "Copepod" = "pink",
               "Diptera" = "hotpink",
               "Hemiptera" = "coral",
               "Nematode" = "aquamarine",
               "Ostracod" = "maroon")) +
# changing default theme
  theme_bw() +
# removing legend
  theme(legend.position = "none")
```

## d. Interpret your figure

All taxon groups seem to have peaks in abundance when salinity is measured at 0, as well as quite a few groups having peaks around a salinity of 40 ppt besides nemotodes and ostracods. From this graph, it looks like copecods have the highest average abundance overall across salinity levels, with Ostracods and Cladoceras following closely behind. This aligns with the idea that Copecod abundance has a stronger relationship to salinity levels compared to all other groups. 


